---
title: "Testing if short time series trends are compatible with long term stable trends"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## *In silico* test

We know short time series are not reliable (refs), and can show transient trends not representative of a medium or long term trend. Here we propose a metric to complement short term linear trend analysis (e.g. using GLM) asking if the observed trend is compatible with a long term trend. This is analogous to a strtuctural stability view, where we test which is the range of short term trends that stable populations can display. Hence, a STS can be declining, but this decline is compatible with stable LTS, or it can be declining and the observed decline is unlikely to come from a stable LTS.

To test this idea we start by simulating a LTS with known parameters. Next, we take a slice of this LTS to crate a STS (years = 6). Finally, we create null models based on the long term parameters, but with a stable trend (slope = 0) and a restricted number of years (years = 6).

```{r}
#source("stability_jga.R")
#source("results_functions.R") #note, need to unify all functions in a single file.
set.seed(1234)
#Create LTS params
n_years <- 20
years <- 1:n_years
r_obs <- -0.02 #slope
phi <- 0 #autocorrelation
sdev <- 0.20 #variability
mu_zero <-  runif(1, 0.2, 90) #intercept

#create a realized time series
LTS <- simulate_ts(r=r_obs, phi=phi, sdev=sdev, mu_zero=mu_zero, years=years, n_rep=1)
plot(LTS$abund ~ LTS$year, t = "b")

#We cut a 6 years STS
#stable
STS_0 <- LTS[8:13,] #manually chossen
STS_D <- LTS[1:6,] #manually chossen
plot(STS_0$abund ~ STS_0$year, t = "b")
plot(STS_D$abund ~ STS_D$year, t = "b")

#See ST trend using GLM
STS_0_model <- run_model(STS_0)
STS_D_model <- run_model(STS_D)
#I ignore warnings for now...

#Create null models with stable trend and 6 years (TS + modles)
null <- simulate_ts(r=0, phi=phi, sdev=sdev, mu_zero=mu_zero, years=1:6, n_rep=100)
null_models <- run_model(null)
#ignoring warnings so far (prune?)


#Finally we need to comapre both
check_results(model_empiric = STS_0_model,
              model_simulated = null_models)
#Here I would phrase it diferently.

check_results(model_empiric = STS_D_model,
              model_simulated = null_models)

plot_results(model_empiric = STS_0_model, 
             model_simulated = null_models)
#check warning

plot_results(model_empiric = STS_D_model, 
             model_simulated = null_models)
#check warning; also check p-val, which does not change.
```

We can say there is only a 20% chances that the 6 year observed trend comes from a stable population. However, note that the STS slope estimated is -0.12, while the LTS trend is -0.02. So this is a SING test, not a magnitude test.

Note, for stable STS trends, we can't say anything. This tool is only to catch transient trends that are compatible with declines. If anything, we would need to build a null model with slope = -0.02 and ask if a stable trend is compatible with a declining LTS.

Also note that for positive slopes we need to do 1-q to obntain the probability. This means a 50% prob is the max we can attain? This is not intuitive :)

## Run multiple simulations

Let's do the same, but for many simulations. We need to store for each LTS, its slope, and the window slices STS trends and p-values and null %.

```{r}
out <- data.frame(LTS_rep = NA, LTS_slope = NA, 
                  STS_rep = NA, STS_slope = NA, STS_s_pv = NA,
                  STS_null = NA)

for(i in 1:100){ #100 replicates
  #create LTS
  n_years <- 20
  years <- 1:n_years
  r_obs <- -0.02 #slope
  phi <- 0 #autocorrelation
  sdev <- 0.20 #variability
  mu_zero <-  runif(1, 0.2, 90) #intercept
  #create a realized time series
  LTS <- simulate_ts(r=r_obs, phi=phi, sdev=sdev, mu_zero=mu_zero,
                     years=years, n_rep=1)
  out[i,"LTS_rep"] <- i
  out[i,"LTS_slope"] <- r_obs #other params can be stored
  #We cut a 6 years STS
  window <- 1:6
  n_models <- n_years-length(window)+1
  STS_chunks <- vector("list", n_models)
  #Create null models with stable trend and e.g. 6 years (TS + modles)
  null <- simulate_ts(r=0, phi=phi, sdev=sdev, mu_zero=mu_zero, 
                        years=window, n_rep=100)
  null_models <- run_model(null)
  #loop through chunks
  for(j in 1:n_models){
    STS_temp <- LTS[window + j - 1, ] 
    STS_chunks[[j]] <- STS_temp
    #See ST trend using GLM
    STS_model_temp <- run_model(STS_temp)
    out[i+j,"STS_rep"] <- j    
    out[i+j,"STS_slope"] <- STS_model_temp$r_estimated
    out[i+j,"STS_s_pv"] <- STS_model_temp$slope_pval
    out[i+j,"STS_null"] <- check_results(model_empiric = STS_model_temp,
                                         model_simulated = null_models)
  }
}
#make bins
out$bins <- ifelse(out$STS_slope < -0.1 & out$LTS_slope = 0.02, "-ST-LT",
                   ifelse(out$STS_slope < -0.1 & out$LTS_slope = 0, 
                          "-ST*LT",
                          ))

boxplot(out$STS_null ~ out$bins)
```
